Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-23 paper/paper.tex
Update paper/paper.tex

In this paper, we aim to gain insight into resource management challenges when executing 
scientific workflow ensembles on clouds. In particular, we address a new and important 
problem of maximizing the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation for this work is to answer the fundamental question of 
concern to a researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble based on the constraints. We evaluate these algorithms using a
simulator we have developed using CloudSim~\cite{Calheiros11}, which models the 
infrastructure and the application. The algorithms were evaluated on a set 
of scientific workflow ensembles from several different applications using a 
broad range of budget and deadline parameters.

@@

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniqies. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware scheduling heuristics that cannot be applied to
independent tasks.

Our work is related to the strategies for deadline-constrained cost-minimization
workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also have budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore cost is not something we optimize, but rather a constraint.

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in practice. Our work can be also
regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume that we have runtime estimates for each task in the workflow based on
either a performance model of the application, or historical data that can be
mined. In this study we use synthetic workflows that were generated using 
historical data from real applications~\cite{Bharathi08}. These applications 
come from a wide variety of domains including: bioinformatics (Epigenomics, 
SIPHT), astronomy (Montage), earthquake science (CyberShake), and physics 
(LIGO).

@@

This algorithm guarantees that tasks from lower priority workflows are
always deferred when higher-priority tasks are available, but lower-priority
tasks can still occupy idle VMs when higher-priority tasks are not available. 
However, because there is no preemption, long-running low-priority tasks may delay 
the execution of higher-priority tasks. In addition, tasks from low priority 
workflows may be executed even though there is no chance that those workflows 
will be completed within the current budget and deadline. Fig.~\ref{fig:spds-example} 
shows an example schedule generated using the SPDS algorithm. The figure illustrates 
how tasks from lower priority workflows backfill idle VMs when tasks from higher 
priority workflows are not available.

@@

An example schedule generated by SPSS is shown in Fig.~\ref{fig:spss-example}.
This example shows how SPSS tends to start many workflows in parallel, running
each workflow over a longer period of time on only a few VMs to minimize cost. 
In comparison, the dynamic algorithms tend to run one workflow at a time across 
many VMs in parallel.

@@

The constant distribution produces ensembles containing of workflows with the same number of tasks per workflow, 
while the Pareto-like distribution produces ensembles with a small number of larger 
workflows and a large number of smaller workflows, as seen in 
Fig.~\ref{fig:ensemble-distribution}. The number of large workflows (of size $\geq$ 900) 
is slightly increased to represent the ``heavy-tail'', i.e. that there is a 
not-insignificant number of large workflows in the ensemble. We assume that large
workflows are more important to users, so in the Pareto ensembles the workflows are 
assigned priorities according to their size so that the largest workflows are given 
the highest priority. In
future work we plan to explore other ensemble types such as uniform distributions.

@@

We show the results for three algorithms: DPDS, WA-DPDS and SPSS. These 
experiments were run with maximum autoscaling factor set to 1.0 for
DPDS and WA-DPDS. After experimenting with DPDS and WA-DPDS we found that, 
due to the high parallelism of workflows, the resource utilization remains 
high most of the time, which leads to the rapid provisioning of new resources 
at the beginning of ensemble execution. This, in turn, results in less 
efficient total utilization due to high levels of parallelism. Setting the 
scaling limit to 1.0, ensures that the shape of the resource provisioning 
plan  is close to a rectangle (see Fig.~\ref{fig:spds-example}) and the VM 
termination procedure (Algorithm~\ref{alg:prov}) allows maximum utilization 
of VMs before shutdown. Based on experiments with different applications, 
we set the $\alpha$ parameter for deadline distribution in SPSS to be 0.7.

@@

Fig.~\ref{fig:number-complete-pareto} shows the number of the workflows
completed for two budget values for each application ensemble. The budget values
selected correspond to two extreme cases: {\em small budget} (top) and {\em
large budget} (bottom). The small budgets permit the algorithms to complete 
only a small number of workflows from the ensemble, while the larger budgets 
enable the algorithms to finish almost all the workflows. The rangesIn three of
five cases the SPSS algorithm performs better for tighter constraints. When the 
budget is low, or the deadline is short, the static procedure is able to more 
efficiently pack workflow tasks onto limited resources. This is the case for 
the LIGO, Epigenomics and SIPHT ensembles for small budgets, and for large
budgets with short deadlines. The exceptions are the Montage and CyberShake 
ensembles,  which consist of many fine-grained tasks. In these applications, 
as well for the other ensembles with large budgets and deadlines, the simple 
dynamic scheduling techniques perform equally well or better. The reason for 
this behavior is that the large number of tasks guarantees that resources 
are almost never idle regardless of how the tasks are scheduled. We also 
observe that WA-DPDS performs better than the workflow-unaware version, DPDS. 
This means that for the workflows we tested the simple admission procedure based 
on estimation of workflow cost does not degrade the solution, but rather in 
many cases it enables larger workflows that would lead to budget overrun to
be rejected. Thus it can save the space for smaller workflows to complete.

%Fig.~\ref{fig:score} shows the score $s$ as defined in Eq.~\ref{eq:score}
%computed for two sample applications. Comparing
%Fig.~\ref{fig:number-complete-pareto} to Fig.~\ref{fig:score} we can see that
%the latter one gives more flattened shape: this means that e.g. the improvement
%of the solution by one workflow may mean adding a one low-priority workflow.
%Nevertheless we can see that the curves on both figures have a very similar
%shape, which mens that our algorithms obey the priority rules and do not try to
%admit lower-priority workflows at the cost of the more important ones.

@@

Fig.~\ref{fig:total-time} presents the same results using the total computing
time metric, $T_C$. These plots confirm our previous observations about the
algorithms' performance, but they also allow us observe a more general property:
for a given budget, when the deadline is tightened the amount of work done
steeply decreases. This is partially due to the fact that fewer workflows can be
completed given a short deadline because their critical paths are simply too long
and they are squeezed out of early start times by high priority workflows.
In addition, assigning a shorter deadline means that more VMs need to be
provisioned to complete the work. However, because of the dependencies in the 
workflows, many of these VMs will either be idle, or will be backfilled with 
tasks from low-priority workflows that never finish, resulting, effectively, 
in reduced resource utilization and fewer completed workflows.

These observations are further confirmed when examining the effective computing
cost, as shown in Fig.~\ref{fig:cost}. It can be seen that lower resource
utilization and lower parallel efficiency for short deadlines lead to
substantial increases in the effective computing cost. This increase can be 
significant. For example, a SIPHT ensemble with short deadlines and a budget
of \$1,000 has an effective computing cost $\sim$8--20\% higher than the 
minimum effective cost of \$1/hour, depending on the algorithm used. This means 
that out of the \$1,000 budget, \$80--\$200 is spent on idle resources.

Another general conclusion can be drawn from Fig.~\ref{fig:cost}: despite some
random fluctuations resulting mostly from the finite result space, all the
curves for all the applications and budgets have a similar shape. This shape
represents a trade-off curve between two conflicting objectives: time and cost.
We observe that the selection of deadline and cost can be formulated as a
multi-criteria optimization problem, and that the curves presented in 
Fig.~\ref{fig:cost} are approximations of the Pareto front (or Pareto set) of
solutions to this problem. In our case the solutions were achieved by maximizing
the amount of work done, which is the inverse of minimizing the effective 
computing cost metric for a given deadline. Therefore, the obtained results 
can be used to aid in resource allocation planning, when both cost and time 
criteria are given not as constraints but as objectives.

@@

Fig.~\ref{fig:budgets} shows the same data for the SIPHT ensemble from another
perspective. It helps answer the question: Given a deadline, how many workflows
can be finished when we increase the budget? In Fig.~\ref{fig:budgets}(a) the
growth is super-linear, which results from the Pareto-like distribution of 
workflow sizes in the ensemble (see Fig.~\ref{fig:ensemble-distribution}). 
For ensembles of constant size workflows (see Fig.~\ref{fig:budgets}(b)) 
the growth is linear and the costs are higher, since all the worklfows in the
test consisted of 1000 tasks.

>>>>

In this paper, we aim to gain insight into resource management challenges when executing 
scientific workflow ensembles on clouds. In particular, we address a new and important 
problem of maximizing the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation for this work is to answer the fundamental question of 
concern to a researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble based on the constraints. We evaluate these algorithms using a
simulator we have developed using CloudSim~\cite{Calheiros11}, which models the 
infrastructure and the application. The simulator includes also tightly-coupled 
scheduler and provisioner modules. The algorithms were evaluated on a set 
of scientific workflow ensembles from several different applications using a 
broad range of budget and deadline parameters.

@@

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniques. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware scheduling heuristics that cannot be applied to
independent tasks.

Our work is related to the strategies for deadline-constrained cost-minimization
workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also consider budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore cost is not something we optimize (i.e. an objective), but rather a constraint.

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in the scenarios we consider in this paper. Our work can be also
regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume that we have runtime estimates for each task in the workflow based on
either a performance model of the application, or historical data that can be
mined. In this study we use synthetic workflows that were generated using 
historical data from real applications~\cite{Bharathi08}. These applications 
come from a wide variety of domains including: bioinformatics (Epigenomics, 
SIPHT: sRNA identification protocol using high-throughput
technology), astronomy (Montage), earthquake science (CyberShake), and physics 
(LIGO).

@@

This algorithm guarantees that tasks from lower priority workflows are
always deferred when higher-priority tasks are available, but lower-priority
tasks can still occupy idle VMs when higher-priority tasks are not available. 
However, because there is no preemption, long-running low-priority tasks may delay 
the execution of higher-priority tasks. In addition, tasks from low priority 
workflows may be executed even though there is no chance that those workflows 
will be completed within the current budget and deadline. Figure~\ref{fig:spds-example} 
shows an example schedule generated using the SPDS algorithm. The figure illustrates 
how tasks from lower priority workflows backfill idle VMs when tasks from higher 
priority workflows are not available.

@@

An example schedule generated by SPSS is shown in Figure.~\ref{fig:spss-example}.
This example shows how SPSS tends to start many workflows in parallel, running
each workflow over a longer period of time on only a few VMs to minimize cost. 
In comparison, the dynamic algorithms tend to run one workflow at a time across 
many VMs in parallel.

@@

The constant distribution produces ensembles containing of workflows with the same number of tasks per workflow, 
while the Pareto-like distribution produces ensembles with a small number of larger 
workflows and a large number of smaller workflows, as seen in 
Figure~\ref{fig:ensemble-distribution}. The number of large workflows (of size $\geq$ 900) 
is slightly increased to represent the ``heavy-tail'', i.e. that there is a 
not-insignificant number of large workflows in the ensemble. We assume that large
workflows are more important to users, so in the Pareto ensembles the workflows are 
assigned priorities according to their size so that the largest workflows are given 
the highest priority. In
future work we plan to explore other ensemble types such as uniform distributions.

@@

We show the results for three algorithms: DPDS, WA-DPDS and SPSS. These 
experiments were run with maximum autoscaling factor set to 1.0 for
DPDS and WA-DPDS. After experimenting with DPDS and WA-DPDS we found that, 
due to the high parallelism of workflows, the resource utilization remains 
high most of the time, which leads to the rapid provisioning of new resources 
at the beginning of ensemble execution. This, in turn, results in less 
efficient total utilization due to high levels of parallelism. Setting the 
scaling limit to 1.0, ensures that the shape of the resource provisioning 
plan  is close to a rectangle (see Figure~\ref{fig:spds-example}) and the VM 
termination procedure (Algorithm~\ref{alg:prov}) allows maximum utilization 
of VMs before shutdown. Based on experiments with different applications, 
we set the $\alpha$ parameter for deadline distribution in SPSS to be 0.7.

@@

Figure~\ref{fig:number-complete-pareto} shows the number of the workflows
completed for two budget values for each application ensemble. The budget values
selected correspond to two extreme cases: {\em small budget} (top) and {\em
large budget} (bottom). The small budgets permit the algorithms to complete 
only a small number of workflows from the ensemble, while the larger budgets 
enable the algorithms to finish almost all the workflows. The rangesIn three of
five cases the SPSS algorithm performs better for tighter constraints. When the 
budget is low, or the deadline is short, the static procedure is able to more 
efficiently pack workflow tasks onto limited resources. This is the case for 
the LIGO, Epigenomics and SIPHT ensembles for small budgets, and for large
budgets with short deadlines. The exceptions are the Montage and CyberShake 
ensembles,  which consist of many fine-grained tasks. In these applications, 
as well for the other ensembles with large budgets and deadlines, the simple 
dynamic scheduling techniques perform equally well or better. The reason for 
this behavior is that the large number of tasks guarantees that resources 
are almost never idle regardless of how the tasks are scheduled. We also 
observe that WA-DPDS performs better than the workflow-unaware version, DPDS. 
This means that for the workflows we tested the simple admission procedure based 
on estimation of workflow cost does not degrade the solution, but rather in 
many cases it enables larger workflows that would lead to budget overrun to
be rejected. Thus it can save the space for smaller workflows to complete.

%Figure~\ref{fig:score} shows the score $s$ as defined in Eq.~\ref{eq:score}
%computed for two sample applications. Comparing
%Figure~\ref{fig:number-complete-pareto} to Figure~\ref{fig:score} we can see that
%the latter one gives more flattened shape: this means that e.g. the improvement
%of the solution by one workflow may mean adding a one low-priority workflow.
%Nevertheless we can see that the curves on both figures have a very similar
%shape, which mens that our algorithms obey the priority rules and do not try to
%admit lower-priority workflows at the cost of the more important ones.

@@

Figure~\ref{fig:total-time} presents the same results using the total computing
time metric, $T_C$. These plots confirm our previous observations about the
algorithms' performance, but they also allow us observe a more general property:
for a given budget, when the deadline is tightened the amount of work done
steeply decreases. This is partially due to the fact that fewer workflows can be
completed given a short deadline because their critical paths are simply too long
and they are squeezed out of early start times by high priority workflows.
In addition, assigning a shorter deadline means that more VMs need to be
provisioned to complete the work. However, because of the dependencies in the 
workflows, many of these VMs will either be idle, or will be backfilled with 
tasks from low-priority workflows that never finish, resulting, effectively, 
in reduced resource utilization and fewer completed workflows.

These observations are further confirmed when examining the effective computing
cost, as shown in Figure~\ref{fig:cost}. It can be seen that lower resource
utilization and lower parallel efficiency for short deadlines lead to
substantial increases in the effective computing cost. This increase can be 
significant. For example, a SIPHT ensemble with short deadlines and a budget
of \$1,000 has an effective computing cost $\sim$8--20\% higher than the 
minimum effective cost of \$1/hour, depending on the algorithm used. This means 
that out of the \$1,000 budget, \$80--\$200 is spent on idle resources.

Another general conclusion can be drawn from Figure~\ref{fig:cost}: despite some
random fluctuations resulting mostly from the finite result space, all the
curves for all the applications and budgets have a similar shape. This shape
represents a trade-off curve between two conflicting objectives: time and cost.
We observe that the selection of deadline and cost can be formulated as a
multi-criteria optimization problem, and that the curves presented in 
Figure~\ref{fig:cost} are approximations of the Pareto front (or Pareto set) of
solutions to this problem. In our case the solutions were achieved by maximizing
the amount of work done, which is the inverse of minimizing the effective 
computing cost metric for a given deadline. Therefore, the obtained results 
can be used to aid in resource allocation planning, when both cost and time 
criteria are given not as constraints but as objectives.

@@

Figure~\ref{fig:budgets} shows the same data for the SIPHT ensemble from another
perspective. It helps answer the question: Given a deadline, how many workflows
can be finished when we increase the budget? In Figure~\ref{fig:budgets}(a) the
growth is super-linear, which results from the Pareto-like distribution of 
workflow sizes in the ensemble (see Figure~\ref{fig:ensemble-distribution}). 
For ensembles of constant size workflows (see Figure~\ref{fig:budgets}(b)) 
the growth is linear and the costs are higher, since all the workflows in the
test consisted of 1000 tasks.
